---
title: "ENVS 193DS Final"
author: "Kierra Miller"
date: "2026-06-07"
format:
  pdf:
    toc: true
---

GitHub Repository: <https://github.com/kierramiller/ENVS-193DS_spring-2026_final>




```{r}

# Set up: load packages and dataset 

library(tidyverse)
library(janitor)
library(readxl)
library(DHARMa)
library(AICcmodavg)
library(ggeffects)

nests <- read_csv("nest_data_final.csv")
```

# Problem 1. 

## a. Transparent statistical methods 

In part 1, my co-worker used a logistic regression to evaluate whether distance to the water's edge predicted the probability of American Avocet microhabitat use. In part 2, my co-worker used a chi-square test of independence to evaluate whether habitat use differed among bird species.

## b. Suggestions for rewriting 

Part 1: American Avocets were more likely to use some areas than others based on their proximity to the water's edge, suggesting that distance from water may influence microhabitat selection. 

Part 2: American Avocets, Forster's Terns, and Black-necked stilts used habitat types differently, indicating that each species may have distinct habitat preferences within the San Francisco Bay wetland system. 

## c. Further analysis 

One additional variable that would be beneficial to measure and that could influence the probability of American Avocet microhabitat use is water depth, this could be measured in (cm) at each observation location using a depth gauge or meter stick. By collecting data on water depth you can determine if water depth may influence habitat use because American Avocets typically forage in shallow water, the deeper depths may be too challenging or too scarce on feeding opportunities. 

# Problem 2. Data Analysis 

## a. Clean and wrangle 

```{r}
# Create cleaned dataset containing only variables needed
nests_clean <- nests %>%

  # Keep only columns needed for analysis
  select(
    case_control,
    height_cm,
    ant_species
  ) %>%

  # Keep only the three Crematogaster ant species
  filter(
    ant_species %in% c(
      "AB",
      "RRB",
      "BBR"
    )
  ) %>%

  # Reorder factor levels as instructed
  mutate(
    ant_species = factor(
      ant_species,
      levels = c(
        "AB",
        "RRB",
        "BBR"
      )
    ),

    # Add full scientific names
    ant_species_full = case_when(
      ant_species == "AB" ~ "Crematogaster sjostedti",
      ant_species == "RRB" ~ "Crematogaster mimosae",
      ant_species == "BBR" ~ "Crematogaster nigriceps"
    )
  )

# Display structure
str(nests_clean)

# Display 10 random observations
slice_sample(
  nests_clean,
  n = 10
)
```

## b. Response variable 

The response variable is `case_control`. A value of 1 indicates that a bird nest was present in a tree, whereas a value of 0 indicates that the tree was selected as an available comparison tree without a nest. Biologically, this variable represents whether a tree was occupied as a nesting site by birds.

## c. Purpose of study 

Tree height is a continuous predictor measured in centimeters. As tree height increases, the probability of nest occurrence may increase because taller trees can provide birds with improved visibility, greater protection from some predators, and more attractive nesting locations. This idea is discussed in the Introduction of Lujan et al. (2023), where the authors describe factors influencing nest-site selection and interactions between birds and ant-defended acacia trees.

Ant species is a categorical predictor with three levels: Crematogaster sjostedti, Crematogaster mimosae, and Crematogaster nigriceps. The probability of nest occurrence may differ among ant species because these ants vary in their defensive behaviors and interactions with birds, potentially making some trees safer or more attractive nesting locations than others. This information is discussed in the Introduction and Methods sections describing bird nesting behavior and ant–bird interactions.

## d. Table of models 

| Model Number | Response Variable | Predictors | Interaction? |
|:---|:---|:---|:---|
| 1 | Nest occurrence | Tree height + ant species | No |
| 2 | Nest occurrence | Tree height + ant species | Yes (tree height × ant species) |

## e. Run the models 
```{r}
# Model 1: additive model
m1 <- glm(
  case_control ~ height_cm + ant_species,
  data = nests_clean,
  family = binomial
)

# Model 2: interaction model
m2 <- glm(
  case_control ~ height_cm * ant_species,
  data = nests_clean,
  family = binomial
)

```

## f. Select the best model 
```{r}
# Compare AIC values to select which model is best 
AIC(m1, m2)
```

The best model as determined by Akaike's Information Criterion (AIC) included tree height, ant species, and the interaction between tree height and ant species as predictors of nest occurrence. This model had a lower AIC value (237.8042) than the model containing only the main effects of tree height and ant species (258.7430). 

## g. Check the diagnostics 
```{r}
# Simulate residuals
sim_res <- simulateResiduals(m2)

# Diagnostic plots
plot(sim_res)
```
Diagnostic plots generated using simulated residuals from the DHARMa package indicated that the selected logistic regression model adequately fit the data. The residuals closely followed the expected distribution, and there was no evidence of overdispersion or influential outliers. Overall, no major violations of model assumptions were detected.

## h. Visualize the model predictions 
```{r}
# Generate predicted probabilities from selected model
pred <- ggpredict(
  m2,
  terms = c(
    "height_cm",
    "ant_species"
  )
)

# View prediction object
head(pred)

# Create figure showing model predictions and observed data
ggplot() +

  # Observed data
  geom_jitter(
    data = nests_clean,
    aes(
      x = height_cm,
      y = case_control
    ),
    width = 0,
    height = 0.03,
    alpha = 0.35,
    color = "gray40"
  ) +

  # Confidence interval
  geom_ribbon(
    data = pred,
    aes(
      x = x,
      ymin = conf.low,
      ymax = conf.high
    ),
    fill = "lightblue",
    alpha = 0.3
  ) +

  # Prediction line
  geom_line(
    data = pred,
    aes(
      x = x,
      y = predicted
    ),
    color = "darkblue",
    linewidth = 1.2
  ) +

  # Separate panel for each ant species
  facet_wrap(~group) +

  labs(
    x = "Tree height (cm)",
    y = "Predicted probability of nest occurrence"
  ) +

  theme_classic() +

  theme(
    legend.position = "none"
  )
```


## i. Figure caption 
Figure 1. Predicted probability of bird nest occurrence as a function of tree height across three acacia ant species. Points represent observed nest occurrence data (0 = no nest, 1 = nest present). Solid lines represent predicted probabilities from the selected logistic regression model, and shaded regions represent 95% confidence intervals. Separate panels display model predictions for each ant species. Data from Lujan et al. (2023).

## j. Calculate model predictions 
```{r}
# Predict nest occurrence at 600 cm for each ant species
predictions_600 <- data.frame(
  ant_species = c(
    "Crematogaster sjostedti",
    "Crematogaster mimosae",
    "Crematogaster nigriceps"
  ),
  predicted_probability = pred_600
)

predictions_600
```


## k. Interpret your results 

The probability of nest occurrence differed among acacia ant species and changed with tree height (Figure 1). At the tallest tree height in the dataset (600 cm), the predicted probability of nest occurrence was approximately 0.21 for trees occupied by *Crematogaster sjostedti*, 0.74 for trees occupied by *Crematogaster mimosae*, and 1.00 for trees occupied by *Crematogaster nigriceps*. Figure 1 shows that the relationship between tree height and nest occurrence differed among ant species, which is consistent with the interaction model being selected as the best-supported model by AIC. These patterns suggest that both tree height and ant species may influence nesting decisions by birds. The differences among ant species may occur because ant species vary in their ecological interactions with birds and behaviors, making some trees safer or more suitable nesting locations than others.


# Problem 3. Affective and exploratory visualizations 


##  a. Comparing visualizations 

## b. Designing an analysis 


## c. Check any assumptions and run your analysis 

## d. Create a visualization 

## e. Write a caption 
## f. Write about your results 
## g. Sharing your affective visualization 







